Simple repression draft

### 2020-12-06 doc/resubmission/section_01_introduction.tex
minor fixes to typos
minor changes after 2nd round of reviews. Paper is ready

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site as developed in pioneering early work on the quantitative
dissection of transcription \cite{Oehler1994, Oehler1990}. One of the main
features of the models we explore is that, by construction, all features related
to the microstates in which the repressor is bound to the promoter can be
separated from the microstates in which the RNA polymerase (RNAP) is bound. From
a modeling perspective, this means that some of the models can be written as
effective two-state models for which there is a rich
literature~\cite{Peccoud1995, Shahrezaei2008, Iyer-Biswas2009, Tkacik2009,
Sanchez2013, Jones2014, So2011, Munsky2012}. Here, we systematically compare the
predictions of several models with different levels of coarse graining written
in terms of thermodynamic and kinetic parameters. One goal in exploring such
coarse-grainings is to build towards the future models of regulatory response
that will be able to serve the powerful predictive role needed to take synthetic
biology from a brilliant exercise in enlightened empiricism to a rational design
framework as in any other branch of engineering. More precisely, we want
phenomenology in the sense of coarse-graining away atomistic detail, but still
retaining biophysical meaning. In particular a key question is: at this level of
coarse-graining, what microscopic details do we need to explicitly model, and
how do we figure that out? For example, do we need to worry about all or even
any of the steps that individual RNA polymerases go through each time they make
a transcript? Turning the question around, can we see any imprint of those
processes in the available data? If the answer is no, then those processes are
irrelevant for our purposes. Forward modeling and inverse (statistical
inferential) modeling are necessary to tackle such questions. We combine both
approaches in order to discard models that cannot empirically satisfy the main
features of experimental data. First we apply forward modeling to demonstrate
that none of the models are distinguishable at the level of mean gene
expression. We then extend the modeling to look at higher moments of the
distribution, eliminating models that do not empirically satisfy the observed
cell-to-cell variability. Finally we arrive at a minimal model on which we can
apply inverse modeling in order to infer the parameters that explain the data.

@@

\afterpage{\clearpage}
\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{../../figures/main/fig01.pdf}
\caption{\textbf{An overview of the simple repression motif at the level of
means.} (A) Schematic of the qualitative biological picture of the simple
repression genetic architecture. (B) and (C) A variety of possible
mathematicized cartoons of simple repression, along with the effective parameter
$\rho$ which subsumes all regulatory details of the architecture that do not
directly involve the repressor. (B) Simple repression models from an equilibrium
perspective. (C) Equivalent models cast in chemical kinetics language. (D) The
``master curve'' to which all cartoons in (B) and (C) collapse.}
\label{fig1:means_cartoons}
\end{figure}

>>>>

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site as developed in pioneering early work on the quantitative
dissection of transcription \cite{Oehler1994, Oehler1990}. One of the main
features of the models we explore is that, by construction, all features related
to the microstates in which the repressor is bound to the promoter can be
separated from the microstates in which the RNA polymerase (RNAP) is bound. From
a modeling perspective, this means that some of the models can be written as
effective two-state models for which there is a rich
literature~\cite{Peccoud1995, Raj2006, Shahrezaei2008, Iyer-Biswas2009,
Tkacik2009, Sanchez2013, Jones2014, So2011, Munsky2012, Kumar2015}. Here, we
systematically compare the predictions of several models with different levels
of coarse graining written in terms of thermodynamic and kinetic parameters. One
goal in exploring such coarse-grainings is to build towards the future models of
regulatory response that will be able to serve the powerful predictive role
needed to take synthetic biology from a brilliant exercise in enlightened
empiricism to a rational design framework as in any other branch of engineering.
More precisely, we want phenomenology in the sense of coarse-graining away
atomistic detail, but still retaining biophysical meaning. In particular a key
question is: at this level of coarse-graining, what microscopic details do we
need to explicitly model, and how do we figure that out? For example, do we need
to worry about all or even any of the steps that individual RNA polymerases go
through each time they make a transcript? Turning the question around, can we
see any imprint of those processes in the available data? If the answer is no,
then those processes are irrelevant for our purposes. Forward modeling and
inverse (statistical inferential) modeling are necessary to tackle such
questions. We combine both approaches in order to discard models that cannot
empirically satisfy the main features of experimental data. First we apply
forward modeling to demonstrate that none of the models are distinguishable at
the level of mean gene expression. We then extend the modeling to look at higher
moments of the distribution, eliminating models that do not empirically satisfy
the observed cell-to-cell variability. Finally we arrive at a minimal model on
which we can apply inverse modeling in order to infer the parameters that
explain the data.

@@

\afterpage{\clearpage}
\begin{figure}[p]
\centering
\includegraphics[width=\textwidth]{../../figures/main/fig01.pdf}
\caption{\textbf{An overview of the simple repression motif at the level of
means.} (A) Schematic of the qualitative biological picture of the simple
repression genetic architecture. (B) and (C) A variety of possible
mathematicized cartoons of simple repression, along with the effective parameter
$\rho$ which subsumes all regulatory details of the architecture that do not
directly involve the repressor. (B) Simple repression models from a
thermodynamic perspective. (C) Equivalent models cast in chemical kinetics
language. (D) The ``master curve'' to which all cartoons in (B) and (C)
collapse.}
\label{fig1:means_cartoons}
\end{figure}
